# Introduction to Statistics

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE)
```

In this lesson you will be introduced to the process of conducting statistical tests in R, specifically chi-square, t-tests, and correlation tests.

First, to access the dataset(s) you will be using today install the `remotes` package, and then install the `lterdatasampler` package (`remotes` is needed because `lterdatasampler` has to be installed from Git Hub as opposed to CRAN).

```{r eval=FALSE}
install.packages("remotes")
remotes::install_github("lter/lterdatasampler")
```

Now load in all libraries needed for this lesson:

```{r}
library(tidyverse)
library(lterdatasampler)
```

Then run the following line of code to retrieve the `and_vertebrates` data set and bring it into your R session:

```{r}
data(and_vertebrates)
```

## Explore the dataset

Do a little exploration of this data first to understand its structure, variables and data types:

```{r}
# View the data structure
glimpse(and_vertebrates)

# Explore the metadata in the Help pane
?and_vertebrates
```

This data set contains length and weight observations for three aquatic species in clear cut and old growth coniferous forest sections of Mack Creek in HJ Andrews Experimental Forest in Oregon. The three species are Cutthroat trout, Coastal giant salamander and Cascade torrent salamander.

## Chi-square - Categorical Analysis

When you are working with two categorical variables, the statistical test you would use is a Chi-square test. This test can tell you if there is a relationship between your two categorical variables.

For example, we have two categorical variables in the `and_vertebrates` data set:

-   `section` = two forest sections, clear cut (CC) and old growth (OG)

-   `unittype` = channel unit classification type (C = cascade, I = riffle, IP = isolated pool (not connected to channel), P = pool, R = rapid, S = step (small falls), SC = side channel, NA = not sampled by unit)

Lets focus this question on cutthroat trout. First explore the abundance of cutthroat trout in different channel types.

```{r}
and_vertebrates %>% 
  filter(species == "Cutthroat trout") %>% 
  group_by(unittype) %>% 
  summarise(abundance = n())
```

This output tells us that there are quite a few observations with the `NA` category, meaning channel types was unknown. Let's edit the workflow above slightly, using two new functions `drop_na()` and `count()`. `drop_na()` will remove any rows within a specified column (or columns) that have NA values, and we can use `count()` as an alternative to `group_by()` and `summarise()` when we just want number of observations for a single variable (in this case `unittype`).

```{r}
and_vertebrates %>% 
  filter(species == "Cutthroat trout") %>% 
  drop_na(unittype) %>% 
  count(unittype)
```

This returns just about the same data frame as the first method, but now with the NA category removed.

From this we also see the highest Cutthroat trout abundances are found in cascade (C), pool (P), and side channel (SC) habitats.

Now, our question expands beyond this one categorical variable (channel type) and we want to know if abundance is affected by both channel and and forest type (`section`). Here, our null hypothesis is that forest and channel type are independent. To test this, we use the `chisq.test()` , but first we have to reformat our data into a **contingency table**. A contingency table is in matrix format, where each cell is the frequency (in this case seen as abundance) of Cutthroat trout in each combination of categorical variables (forest type and channel unit). We can create a contingency table with the `table()` function.

```{r}
# First clean the dataset to create the contingency table from
trout_clean <- and_vertebrates %>% 
  #filter out Cutthroat trout
  filter(species == "Cutthroat trout") %>% 
  # lets test using just the 3 most abundant unittypes
  filter(unittype %in% c("C", "P", "SC")) %>% 
  # drop NAs for both unittype and section
  drop_na(unittype, section)


cont_table <- table(trout_clean$section, trout_clean$unittype)
```

To execute the Chi Square test does not take that much code, but it is important to note that by default, `chisq.test()` assumes the null hypothesis is that all frequencies have equal probability. If you have different pre-conceived frequency probabilities for your data you have to define those within the `chisq.test()` function.

```{r}
chisq.test(cont_table)

```

Lets look at the abundance distribution visually:

```{r}
trout_clean %>% 
  count(unittype, section) %>% 
  ggplot(aes(x = unittype, y = n))+
  geom_col(aes(fill = section))+
  scale_fill_manual(values = c("orange", "darkgreen"))+
  theme_minimal()
```

## t-test - Compare two means

Previous work has shown that forest harvesting can impact aquatic vertebrate biomass (Kaylor & Warren 2017). With this `and_vertebrates` data set we can investigate this hypothesis, by comparing weight to forest type (clear cut or old growth). This therefore involves a test comparing the means (average weight) among two groups (clear cut and old growth forests), which then requires a t-test.

Lets focus on conducting this test for just Cutthroat trout (to reduce species-level variances in weight), so we can use the same `trout_clean` dataset we made earlier, but let's also drop all NAs in `weight_g`. Then, lets first visualize the differences in weight among forest type with a boxplot:

```{r}
trout_clean %>% 
  drop_na(weight_g) %>% 
  ggplot(aes(x = section, y = weight_g))+
  geom_boxplot()
```

We don't see too much of a difference based on this visual, but lets conduct the statistical test to see if your hypothesis is supported.

We can do this with the `t.test()` function in R by specifying your dependent (in our case trout weight) and independent (forest type) variables, which are entered in the `t.test()` function as vectors (instead of just column names like you can do in the Tidyverse). Remember how we can index single columns of data frames with the `$` operator. The order of the variables in the `t.test()` function is {dependent variable} \~ {independent variable}.

```{r}

t.test(trout_clean$weight_g ~ trout_clean$section)
```

The output of this test gives us the test statistics, p-value, and the means for each of our forest groups. Given the very small p-value, we can conclude that we reject the null hypothesis that mean Cutthroat weight is the same in clear cut and old growth forest sections, and looking at our results (specifically the means) we can conclude that Cutthroat trout weight was observed to be significantly higher in clear cut forests compared to old growth forests.

How does this relate to your original hypothesis?

::: {.alert .alert-info}
Note: In the `t.test()` function you can add `paired = TRUE` to conduct a paired t-test. These are for cases when the groups are 'paired' for each observation, meaning each group/treatment was applied to the same individual, such as before and after experiments.
:::

## Correlation - Assess relationships

When you want to assess the relationship between two **continuous variables**, the test you would use is a correlation test. Correlation tests asses both the presence of a significant relationship along with the strength of that relationship (i.e., the correlation coefficient).

For our `and_vertebrates` data set, we can test length-mass relationships for our species with our length and weight continuous variables. Lets test the hypothesis that body length is positively correlated with weight, such that longer individuals will also weight more, specifically looking at the Coastal Giant salamander.

First let's clean our data set to just include the Coastal giant salamander and remove missing values for length and weight. Let's focus on the variable 'length_2\_mm' for SVL (snout-vent length).

```{r}
sally_clean <- and_vertebrates %>% 
  filter(species == "Coastal giant salamander") %>% 
  drop_na(length_2_mm, weight_g)
  
```

Now we can perform the correlation test with the `cor.test()` function. There are multiple correlation methods you can use with this function, by default it uses the Pearson correlation method. However this test assumes that your data is normally distributed, so if that is not the case you can specify `spearman` for `method =` to use a Spearman Rank correlation test, a non-parametric test that is not sensitive to the variable distribution.

Let's look at the distribution of these variables first:

```{r}
hist(sally_clean$length_2_mm)

hist(sally_clean$weight_g)
```

They both look pretty left skewed, therefore likely not normally distributed. We can statistically test if a variable fits a normal distribution with the `shapiro.test()` function. However note that this function only runs for 5000 observations or less, so lets test for normally of the first 5000 obs of our `sally_clean` data set:

```{r}
shapiro.test(sally_clean$length_2_mm[1:5000])

shapiro.test(sally_clean$weight_g[1:5000])
```

The null hypothesis of the Shapiro-Wilk normality test is that the variable is normally distirbuted, so a significant p-value less than 0.05 (as we see for both of our variables here) tells use that our data does not fit a normal distribution.

Therefore we have two options:

1) transform the data or 2) use the non-parametric Spearman correlation test instead of the Pearson's correlation test.

Lets try the first option by log transforming our variables, first viewing the new distribution and then performing the Pearson's correlation test (the default for `cor.test()`).

```{r}
hist(log(sally_clean$length_2_mm))

hist(log(sally_clean$weight_g))
```

Now try the Pearson's test. All we need to add to the `cor.test()` argument is the two variables of our `sally_clean` data set we want to test a relationship for, and let's keep them log-transformed since those distributions looked closer to a normal distribution (visually at least).

```{r}
cor.test(log(sally_clean$length_2_mm), log(sally_clean$weight_g))
```

Okay, from these results we see a very small p-value, meaning there is a significant association between the two, and a correlation coefficient of 0.98, representing a very strong, positive correlation.

Let's look at this correlation visually:

```{r}
sally_clean %>% 
  ggplot(aes(x = length_2_mm, y = weight_g))+
  geom_point()
```

We can use `geom_smooth()` to add a line of best fit using a linear model equation (which you will learn more about next week)

```{r}
sally_clean %>% 
  ggplot(aes(x = length_2_mm, y = weight_g))+
  geom_point()+
  geom_smooth()
```

This tells us it used the 'gam' method, and this best-fit relationship shows a non-linear relationship.

Let's now perform the correlation test again but with `method = 'spearman'`, as the Spearman test is better for non-parametric and non-linear data sets. Therefore we do not need to transform our variables this time.

```{r}
cor.test(sally_clean$length_2_mm, sally_clean$weight_g)
```

These results still represent a significant, positive relationship between SVL and weight for the Coastal Giant salamander, but this test gives a slightly less correlation coefficient compared to the Pearson's test.

```{r}
and_vertebrates %>% 
  filter(species != "Cascade torrent salamander" & !is.na(species)) %>% 
  count(species, sitecode) %>% 
  ggplot(aes(x = sitecode, y = n))+
  geom_col(aes(fill = species))

```

## Exercises

Each question requires you to carry out a statistical analysis to test some hypothesis related to the `and_vertebrates` data set. To answer each question fully:

-   Include the code you used to clean the data and conduct the appropriate statistical test.

-   Report the findings of your test in proper scientific format (with the p-value in parentheses).

1.  Conduct a chi-square test similar to the one we carried out earlier in this lesson plan, but test for a relationship between forest type (`section`) and channel unit (`unittype`) for Coastal giant salamander abundance. Keep all unittypes instead of filtering any like we did for the Cutthroat trout (9 pts.)

2.   Test the hypothesis that there is a significant difference in species biomass between clear cut and old growth forest types for the Coastal Giant salamander. (Note: the steps to carry out this test are similar to what you did in this lesson plan for the Cutthroat trout) (8 pts.)

3.  

### Acknowledgements

Thanks to the developers of [`lterdatasampler`](https://lter.github.io/lterdatasampler/index.html) for providing the data set and vignettes that helped guide the creation of this lesson plan.

### Citations

***Data Source:*** Gregory, S.V. and I. Arismendi. 2020. Aquatic Vertebrate Population Study in Mack Creek, Andrews Experimental Forest, 1987 to present ver 14. Environmental Data Initiative. <https://doi.org/10.6073/pasta/7c78d662e847cdbe33584add8f809165>

Kaylor, M.J. and D.R. Warren. 2017. Linking riparian shade and the legacies of forest management to fish and vertebrate biomass in forested streams. Ecosphere *8*(6). <https://doi.org/10.1002/ecs2.1845>
