# PCA demo

```{r}
library(tidyverse)
library(lterdatasampler)
library(ggfortify)

data("pie_crab")
```


Check correlations:
```{r}
 pie_crab %>%
  #reduce to quant variables
  select(latitude, air_temp:water_temp_sd) %>% 
  cor()
```

```{r}
crab_pca <- pie_crab %>%
  #reduce to quant variables
  select(latitude, air_temp:water_temp_sd) %>%
  prcomp(scale = TRUE)
```


```{r}
summary(crab_pca)
```

```{r}
crab_pca$rotation
```

```{r}
screeplot(crab_pca)
```



```{r}
biplot(crab_pca)
```


Make this look better with ggfortify (and more customization) then put in ggplotly()

```{r}
autoplot(crab_pca, loadings = TRUE, loadings.label = TRUE)+
  theme_minimal()
```

Add the two PC Axes to original dataset (each axis is orthogonal/opposite which is why we often use the first 2)
PCA used in regression as a method of getting rid of collinearity
show them cor(crab_pac$x)

```{r}
bind_cols(pic_crab, crab_pca$x)

```

Lets do a multiple linear regression w/ these two new axes

```{r}
model <- lm(size ~ PC1 + PC2, data = pie_crab)

summary(model)
```

Look back at the loadings to see which ones most strongly define each axis
```{r}
crab_pca$rotation
```


